Data Exploration: IUCN and AquaMaps

### 2015-11-14 data_explore.Rmd
still updating to get to the point where I can do that damn graph: percent overlap vs area ratio.

#species list 
spp_list <- read.csv('shiny/data/spp_am_v_iucn.csv', stringsAsFactors = FALSE)

@@

### Species Map Function ###
# This function takes a single species scientific name as input, then grabs all 
# occurrence cells and associated Aquamaps probability and/or IUCN proportional area
# per cell
get_spp_map <- function(species){
  spp_id <- spp_list %>%
    filter(sciname == species & is.na(parent_sid)) %>%
    dplyr::select(am_sid, iucn_sid, sciname) %>%
    unique()
  iucn_spp_map <- iucn_spp_cells %>%
    filter(sciname == species) %>%
    group_by(loiczid) %>%        
    summarize(iucn_area = max(prop_area))
  ### The group_by() and summarize() are to collapse cells with overlapped polygons, fixing the duplicate 'by' issue
  am_spp_map   <- am_spp_cells %>%
    filter(am_sid == spp_id$am_sid) %>%
    rename(am_prob = prob)
  spp_map <- full_join(iucn_spp_map, am_spp_map, by = 'loiczid')%>%
    as.data.frame()

@@

spp_list <- spp_list %>%
              mutate(total_area = NA,
                     am_area    = NA,
                     iucn_area  = NA,
                     perc       = NA,
                     am_perc    = NA,
                     iucn_perc  = NA,
                     sm_perc    = NA, #filled with % of smaller range that is within the larger range
                     sm_range   = NA) #what range is smaller

@@

###get just the aquamaps cells
   am_spp_map <- spp_map%>%
                   filter(!is.na(am_prob))%>%
                    dplyr::select(loiczid, am_sid, am_prob)%>% #add in am_prob when we want to look at probability
                     unique()

@@

  spp_list[i, ]$sm_range <- ifelse(length(cells_am)<length(cells_iucn), 'AM', 'IUCN') #what dataset has the smaller range
  spp_list[i, ]$sm_perc <- length(cells_overlap)/min(length(cells_am), length(cells_iucn)) * 100 #this is the total percent of smaller range encompassed in the larger range
cat(sprintf('%s: For species %s, %.2f%% of both datasets overlap, %.2f%% is just AquaMaps and %.2f%% is just IUCN\n', i, spp, perc, perc_am, perc_iucn))

  }

@@

#remove non IUCN species and log total area (for use later on)
spp_list <- spp_list%>%
            filter(spp_group != 'hagfishes' & spp_group != 'non-homalopsids')%>%
             mutate(log_total_area = log(total_area))

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE, warning = FALSE, message = FALSE}

scatter_am_iucn_area <- ggplot(spp_df, aes(iucn_area, am_area, color = iucn_category)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'Area (km2) from IUCN dataset', 
       y = 'Area (km2) from AquaMaps dataset', 
       title = 'Comparing total area of species range maps')

@@

sm_plot <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = "Dataset with smaller range", labels = c("AquaMaps", "IUCN")) + 
  labs(x = 'Percent of smaller range that falls within the larger range', 
       y = 'Proportion of all species (n = 1962)')

@@

print(sm_plot)

@@

reg <- ggplot(spp_list, aes(sm_perc, total_area, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  labs(y = 'Total Area (km2)', 
       x = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

print(reg)

@@

lm_eqn <- function(df){
  m  <- lm(total_area~sm_perc, spp_list)
  eq <- substitute(italic('y') == a + b %.% italic('x')*","~~italic(r)^2~"="~r2, 
        list(a  = format(coef(m)[1], digits = 2), 
             b  = format(coef(m)[2], digits = 2), 
             r2 = format(summary(m)$r.squared, digits = 3)))
  as.character(as.expression(eq))             
}

p1 <- reg + 
  geom_text(x = 40, y = 300000000, label = lm_eqn(spp_list), parse = TRUE)

@@

print(p1)

summary(lm(total_area~sm_perc, spp_list))
```

@@

Perent coverage as function of area, clipping to area below 100,000,000 km^2:
``` {r, echo = FALSE, warning = FALSE, message = FALSE}
# remove all area greater than 100000000

l <- spp_list %>% 
  filter(total_area < 100000000)

r <- ggplot(l, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Total Area (km2)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

print(r)

@@

logA_vs_pct <- ggplot(spp_list, aes(log_total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Log(total Area)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range (log) vs. percent overlap of \nsmaller range with larger')

@@

print(logA_vs_pct)
```

``` {r examine regressions, echo = FALSE, warning = FALSE}

@@

###(5) Examining data broken into quartiles to see if smaller patterns emerge
``` {r examine quartiles, echo = FALSE, warning = FALSE}

@@

###(6) Looking at how many and what type (threat category) of species we are losing by setting a threshold of 0.4 on the aquamaps species.

@@

hist <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', y = 'Proportion of all aquamaps species (17,348 total)')

@@

print(hist)

>>>>

#species list 
spp_list_base <- read.csv('shiny/data/spp_am_v_iucn.csv', stringsAsFactors = FALSE)

@@

### Species Map Function ###
# This function takes a single species scientific name as input, then grabs all 
# occurrence cells and associated Aquamaps probability and/or IUCN proportional area
# per cell
get_spp_map <- function(species){
  spp_id <- spp_list_base %>%
    filter(sciname == species & is.na(parent_sid)) %>%
    dplyr::select(am_sid, iucn_sid, sciname) %>%
    unique()
  iucn_spp_map <- iucn_spp_cells %>%
    filter(sciname == species) %>%
    group_by(loiczid) %>%        
    summarize(iucn_area = max(prop_area))
  ### The group_by() and summarize() are to collapse cells with overlapped polygons, fixing the duplicate 'by' issue
  am_spp_map   <- am_spp_cells %>%
    filter(am_sid == spp_id$am_sid) %>%
    rename(am_prob = prob)
  spp_map <- full_join(iucn_spp_map, am_spp_map, by = 'loiczid')%>%
    as.data.frame()

@@

spp_list <- spp_list_base %>%
              mutate(total_area = NA,
                     am_area    = NA,
                     iucn_area  = NA,
                     perc       = NA,
                     am_perc    = NA,
                     iucn_perc  = NA,
                     sm_perc    = NA, #filled with % of smaller range that is within the larger range
                     sm_range   = NA) #what range is smaller

@@

###get just the aquamaps cells
  am_spp_map <- spp_map %>%
    filter(!is.na(am_prob))%>%
    dplyr::select(loiczid, am_sid, am_prob) %>% #add in am_prob when we want to look at probability
    unique()

@@

  spp_list[i, ]$sm_range <- ifelse(length(cells_am)<length(cells_iucn), 'AM', 'IUCN') #what dataset has the smaller range
  spp_list[i, ]$sm_perc <- length(cells_overlap)/min(length(cells_am), length(cells_iucn)) * 100 
    #this is the total percent of smaller range encompassed in the larger range

  cat(sprintf('%s: For species %s, %.2f%% of both datasets overlap, %.2f%% is just AquaMaps and %.2f%% is just IUCN\n', 
              i, spp, perc, perc_am, perc_iucn))
  }

@@

#remove non IUCN species and log total area (for use later on)
spp_list <- spp_list %>%
            filter(spp_group != 'hagfishes' & spp_group != 'non-homalopsids') %>%
              mutate(log_total_area = log(total_area))

@@

scatter_am_iucn_area <- ggplot(spp_df, aes(x = iucn_area, y = am_area, color = iucn_category)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'Area (km2) from IUCN dataset', 
       y = 'Area (km2) from AquaMaps dataset', 
       title = 'Comparing total area of species range maps')

@@

hist_sm_within_lg <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = "Dataset with smaller range", labels = c("AquaMaps", "IUCN")) + 
  labs(x = 'Percent of smaller range that falls within the larger range', 
       y = 'Proportion of all species (n = 1962)')

@@

print(hist_sm_within_lg)

@@

scatter_area_vs_overlap <- ggplot(spp_list, aes(sm_perc, total_area, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  labs(y = 'Total Area (km2)', 
       x = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

print(scatter_area_vs_overlap)

@@

lm_eqn <- function(df){
  m  <- lm(total_area ~ sm_perc, spp_list)
  eq <- substitute(italic('y') == a + b %.% italic('x')*","~~italic(r)^2~"="~r2, 
        list(a  = format(coef(m)[1], digits = 2), 
             b  = format(coef(m)[2], digits = 2), 
             r2 = format(summary(m)$r.squared, digits = 3)))
  as.character(as.expression(eq))             
}

scatter_area_vs_overlap <- scatter_area_vs_overlap + 
  geom_text(x = 40, y = 300000000, label = lm_eqn(spp_list), parse = TRUE)

@@

print(scatter_area_vs_overlap)

summary(lm(total_area ~ sm_perc, spp_list))
```

@@

Percent coverage as function of area, clipping to area below 100,000,000 km^2:
``` {r, echo = FALSE, warning = FALSE, message = FALSE}
# remove all area greater than 100000000

spp_small_df <- spp_list %>% 
  filter(total_area < 100000000)

scatter_a_vs_overlap_small <- ggplot(spp_small_df, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Total Area (km2)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

print(scatter_a_vs_overlap_small)

@@

scatter_logA_vs_overlap <- ggplot(spp_list, aes(log_total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Log(total Area)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range (log) vs. percent overlap of \nsmaller range with larger')

@@

print(scatter_logA_vs_overlap)
```

``` {r examine regressions, eval = FALSE, echo = FALSE, warning = FALSE}

@@

###(5) Examining data broken into quartiles to see if smaller patterns emerge
``` {r examine quartiles, eval = FALSE, echo = FALSE, warning = FALSE}

@@

###(6) Looking at how many and what type (threat category) of species we are losing by setting a threshold of 0.4 on the aquamaps species

This looks at the full Aquamaps dataset... should it be limited to species with valid categories?

@@

hist_area_vs_threshold <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', y = 'Proportion of all aquamaps species (17,348 total)')

@@

print(hist_area_vs_threshold)
